Computing the bicharacter of huge subspaces of polynomials

### 2019-05-13 article/article.tex
Add content in setion 3

%Tableaux
\newcommand{\diagramme}{\YFrench \Yboxdim{10pt}\yng}
\newcommand{\tableau}{\YFrench \Yboxdim{10pt}\young}
%\newcommand{\diagramme}{\Yboxdim{10pt}\yng}  % notation anglaise
%\newcommand{\tableau}{\Yboxdim{10pt}\young}  % notation anglaise
\newcommand{\petitDiagramme}{\YFrench \Yboxdim{5pt}\yng}
\newcommand{\petitTableau}{\tiny \YFrench \Yboxdim{5pt}\young}
%\newcommand{\petitDiagramme}{\Yboxdim{5pt}\yng}  % notation anglaise
%\newcommand{\petitTableau}{\tiny \Yboxdim{5pt}\young}  % notation anglaise

@@

	\begin{abstract}
		... 
	\end{abstract}

@@

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? } \pauline{introduce inert variables now or later?}
	$$(X_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $X_i$ the i-th set of variables.

@@

	It is well known (\cite{Haiman2002} and \cite{Bergeron2009}) that the harmonics $\harmonics_{n,1}$ are generated by the Vandermonde determinant $V_n$ and closed by partial derivatives. Moreover, the diagonal harmonics $\harmonics_{n,r}$ are generated by $\harmonics_{n,1}$ and closed by the polarization operators. \\

@@

	F. Bergeron recently introduced another generalization of the diagonal harmonics, with inert variables. We add a new row $\theta = (\theta_1, \theta_2, \dots, \theta_n)$ of variables in $X$. The symmetric group still acts diagonally on the rows of $X$, but $GL_r$ now acts only on the rows containing non-inert variables. Like before, we consider the smallest subspace containing a certain determinant and closed by partial derivation and polarization, except that this time the inert variables do not intervene in the operators. The determinant used in this case is defined as follow.

	Let $\lambda$ be a partition of $n$, we also denote $\lambda$ the Ferrer's diagram associated to the partition $\lambda$ and each cell of $\lambda$ is given by the coordinates of its bottom left corner. Then the determinant $V_{\lambda}$ associated to $\lambda$ is given by $$V_{\lambda} = \det (x_i^a \theta_i^b)_{1 \leq i \leq n \atop (a,b) \in \lambda} .$$
	For example, the partition $21$ contains the cells $(0,0)$ and $(1,0)$ in the first row, and the cell $(0,1)$ in the second row, and $V_{21} = -x_2\theta_1 + x_3\theta_1 + x_1\theta_2 - x_3\theta_2 - x_1\theta_3 + x_2\theta_3$. \\

@@

	\begin{definition} \pauline{find the right place}
		The \emph{Steenrod operator} $St_{x_a}^k$ for $k>1$ is given by
		$$St_{x_a}^k = \sum_{i=1}^{n} x_{ai} \partial_{x_{ai}}^k$$
		where $x_a$ is a set of variables of $X$. 
	\end{definition}

@@

	So we took the classical route of realizing the coinvariant ring as the space $\def\harmonics{\mathcal{H}}\harmonics_{k,n}$ of *diagonal harmonic polynomials*.
	This is the space of those polynomials that satisfy the equations given by the multisymmetric functions seen as differential operators.

	Solving those differential equations would not be that tractable either.

	\subsection{Polarization: diagonal harmonic polynomials from harmonic polynomials}

	Define the polarization operators:

	$$P_{i,i'}^d = \sum_j   x_{i',j} \partial_{i,j}^d\,.$$

	For $d=1$, those operators implement the action of $sl_k$ induced by $gl_k$.

@@

	\begin{conjecture}
		Diagonal harmonic polynomials are obtained by polarization of harmonic polynomials in the first set of variables.
	\end{conjecture}

	In the sequel, we assume that the above conjecture holds. The space of harmonic polynomials $\harmonics_n$ itself can be constructed, for example, as the subspace spanned by the VanDerMonde determinant under differentiation.

	In practice, we implemented a general purpose utility in Sage `Space(vectors, operators)` that computes the smallest subspace of a vector space which contains the given vectors and is stable under the action of the given linear operators.

	TODO: put an example here

	No rocket science here: maintain a matrix in row echelon form representing a basis of the space computed so far, and a ToDo list of $op(v)$ that remain to be computed and inserted in the matrix.

	This utility should already have been in Sage, and will end up there.

	Once a basis of $\harmonics_{k,n}$ is constructed, there still remains to recover its bicharacter. For $\GLr$ that's given by the multigrading. For $\Sn$ some additional work has to be done (e.g. computing the traces of the action of permutations of all cycle types).

@@

	Advantages:
	- Writing $N=:\sum N_D$, The complexity becomes $(\sum N_D^2) M$ instead of $(\sum N_D)^2M_D$ (in fact better.
	- The multidegrees, and therefore $\GLr$ character, can be directly read off the result.

@@

	The polarization operators commute with the action of $\Sn$. Therefore,
	we can compute independently each $\Sn$ isotypic component of $\harmonics_{k,n}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics^\lambda\,.$$

@@

	Specht polynomials tend to be big: in $6$ variables, the VanDerMonde determinant has $6!=720$ monomials.

@@

	This trick was already used by François for the polarizations of the VanDerMonde (the diagram space). Using higher Specht polynomials further allowed  to generalize that optimization to each shape.

@@

	The polarization operators skew commute. Therefore it would be in principle sufficient to apply them in some "decreasing order". This sounds promising given that, at this stage about 90\% of the polynomials that are produced are redundant and reduced to zero by the linear algebra.

>>>>

%Tableaux
\newcommand{\diagramme}{\YFrench \Yboxdim{10pt}\yng}
\newcommand{\tableau}{\YFrench \Yboxdim{10pt}\young}
\newcommand{\petitDiagramme}{\YFrench \Yboxdim{5pt}\yng}
\newcommand{\petitTableau}{\tiny \YFrench \Yboxdim{5pt}\young}

@@

	\begin{abstract}
		To complete ... 
	\end{abstract}

@@

	Consider the polynomial ring $P=\QQ[X]$ with $k$ sets of $n$ variables. To fix conventions, we think of $X$ as a matrix of variables: \pauline{small $x$ for variables, capital $X$ for sets of variables? } \pauline{introduce inert variables now or later?}
	$$(X_{i,j})_{0\leq i<k \atop 0\leq j<n}$$
	and we denote by $X_i = (x_{i1}, x_{i2}, \dots, x_{in})$ the i-th set of variables.

@@

	\begin{definition} \pauline{define polarization now or later ?}
		The \emph{polarisation operator} $P_{x_a,x_b}^k$ is given by
		$$P_{x_a,x_b}^k = \sum_{i=1}^{n} x_{bi} \partial_{x_{ai}}^k$$
		where $x_a$ and $x_b$ are two sets of variables of $X$. 
	\end{definition}

	For $=1$, those operators implement the action of $\mathfrak{sl}_r$ induced by $\mathfrak{gl}_r$.

	It is known (\cite{Haiman2002} and \cite{Bergeron2009}) that the harmonics $\harmonics_{n,1}$ are generated by the Vandermonde determinant $V_n$ and closed by partial derivatives. Moreover, the diagonal harmonics $\harmonics_{n,r}$ are generated by $\harmonics_{n,1}$ and closed by the polarization operators.

@@

	F. Bergeron recently introduced another generalization of the diagonal harmonics, with inert variables. We add a new row $\theta = (\theta_1, \theta_2, \dots, \theta_n)$ of variables in $X$. The symmetric group still acts diagonally on the columns of $X$, but $GL_r$ now acts only on the rows containing non-inert variables. Like before, we consider the smallest subspace containing a certain determinant and closed by partial derivation and polarization, except that this time the inert variables do not intervene in the operators. The determinant used in this case is defined as follow.

	Let $\lambda$ be a partition of $n$, we also denote $\lambda$ the Ferrer's diagram associated to the partition $\lambda$ and each cell of $\lambda$ is given by the coordinates $(a,b)$ of its bottom left corner. Then the determinant $V_{\lambda}$ associated to $\lambda$ is given by $$V_{\lambda} = \det (x_i^a \theta_i^b)_{1 \leq i \leq n \atop (a,b) \in \lambda} .$$
	For example, the partition $21$ contains the cells $(0,0)$ and $(1,0)$ in the first row, and the cell $(0,1)$ in the second row, and $V_{21} = -x_2\theta_1 + x_3\theta_1 + x_1\theta_2 - x_3\theta_2 - x_1\theta_3 + x_2\theta_3$. \\

@@

	\begin{definition} \pauline{find the right place for Steenrod op}
		The \emph{Steenrod operator} $St_{x_a}^k$ for $k>1$ is given by
		$$St_{x_a}^k = \sum_{i=1}^{n} x_{ai} \partial_{x_{ai}}^k$$
		where $x_a$ is a set of variables of $X$. 
	\end{definition}

@@

	So we took the classical route of realizing the coinvariant ring as the space $\harmonics_{n,r}$ of diagonal harmonic polynomials.
	This is the space of those polynomials that satisfy the equations given by the multisymmetric functions seen as differential operators.

	Solving those differential equations would not be that tractable either. So instead we used operators (partial derivatives and polarization operators).

	\subsection{Polarization: diagonal harmonics from harmonics}

@@

	\begin{theorem}
		Diagonal harmonic polynomials are obtained by polarization of harmonic polynomials in the first set of variables.
	\end{theorem}

	The space of harmonic polynomials $\harmonics_n$ itself can be constructed, for example, as the subspace spanned by the Vandermonde determinant under differentiation.

	In practice, we implemented a general purpose utility in Sage \texttt{Space(vectors, operators)} that computes the smallest subspace of a vector space which contains the given vectors and is stable under the action of the given linear operators.
	\nicolas{put an example here}

	No rocket science here: maintain a matrix in row echelon form representing a basis of the space computed so far, and a ToDo list of $op(v)$, that remain to be computed and inserted in the matrix, where $op$ is an operator and $v$ an element of the subspace in construction.

	This utility should already have been in Sage, and will end up there. \\

	Once a basis of $\harmonics_{n,r}$ is constructed, there still remains to recover its bicharacter. For $\GLr$ that's given by the multigrading. For $\Sn$ some additional work has to be done (e.g. computing the traces of the action of permutations of all cycle types).

@@

	There are two advantages of doing that. 
	First, writing $N=:\sum N_D$, the complexity becomes $(\sum N_D^2) M$ instead of $(\sum N_D)^2M_D$ (in fact better).
	Second, the multidegrees, and therefore $\GLr$ character, can be directly read off the result.

@@

	The polarization operators, ie the action of $\GLr$, commute with the action of $\Sn$. Therefore, we can compute independently each $\Sn$ isotypic component of $\harmonics_{n,r}$ by starting from a basis of harmonic polynomials that itself is adapted to its decomposition into isotypic components:
	$$\harmonics_{n}=:\bigoplus_\lambda \harmonics^\lambda\,.$$

@@

	Specht polynomials tend to be big: in $6$ variables, the Vandermonde determinant has $6!=720$ monomials.

@@

	This trick was already used by François for the polarizations of the Vandermonde (the diagram space). Using higher Specht polynomials further allowed  to generalize that optimization to each shape.

@@

	The polarization operators skew commute. \pauline{add commutator computation} Therefore it would be in principle sufficient to apply them in some "decreasing order". This sounds promising given that, at this stage about 90\% of the polynomials that are produced are redundant and reduced to zero by the linear algebra.
